Reliable feature point matching

### 2013-08-08 text/thesis.tex
Added a lot of information on spectral matching

		\State $matches_{init} \gets \left(f_i, f_m\right)$

@@

\subsubsection{Experimental setup}
We compare the \emph{MM} and \emph{MMC} algorithms to \emph{Ratio} 
\cite{lowe2004sift} as well as \emph{Isodata} \cite{das2008event} and 
\emph{Spectral} \cite{leordeanu2005spectral}.  \emph{Isodata} and 
\emph{Spectral} uses geometric constraints, whereas \emph{Ratio} does 
not.

@@

\begin{algorithm}
\caption{Isodata Match (\emph{Isodata})}
\label{alg-isodata}
%{\fontsize{10}{10}\selectfont
\begin{algorithmic}
\Require $im_1$ : image, $im_2$ : image, $t \in \mathbb{R}$
\State $matches_{init}\gets \varnothing$
\State $matches_{final}\gets \varnothing$
\State $features_1 \gets getFeatures(im_1)$
\State $features_2 \gets getFeatures(im_2)$
\ForAll{$f_i \in features_1$}
    \State $f_m,f_n \gets get2NNs(f_i, features_2)$
    \State $matches_{init} \gets \left(f_i, f_m\right)$
\EndFor
\State $partitions_1 \gets isodata(getPositions(features_1))$
\State $partitions_2 \gets isodata(getPositions(features_2))$
\State $P_{corr} \gets getMatchMat(partitions_1, partitions_2, 
matches_{init})$
\ForAll{$(i,j) \in indices(P)$}
    \If{$\left(P_{i,j} > 5\right) \wedge \left(P_{i,j} / sum(P_{i}) > 
    0.5\right)$}
        \State $matches_{final} \gets getMatches(matches_{init}, 
        partitions_1, partitions_2, i, j)$
    \EndIf
\EndFor
\Return $matches_{final}$
\end{algorithmic}
%}
\end{algorithm}

The \emph{Isodata} algorithm (Algorithm~\ref{alg-isodata}) 
\cite{das2008event} is based on the idea that local feature points 
within a region in one image often matches to features within a 
corresponding region in the other image. To find these regions the 
algorithm uses the Isodata algorithm \cite{ball1965isodata} to perform 
unsupervised clustering on the feature points given their positions in 
the two images. The Isodata algorithm is similar to K-means with the 
addition that clusters are merged and split during the clustering 
process. The initial set of matches are then filtered based on the 
resulting regions. A match is kept only if it belongs to a region with 
more than 5 matches and where half of the matches link the same two 
regions. $P_{corr}$ in Algorithm~\ref{alg-isodata} is a matrix where 
each row corresponds to a region in $im_1$, each column corresponds to a 
region in $im_2$ and $P_{i,j}$ is the amount of matches going from 
region $i$ to region $j$.

There is no inherent threshold parameter in \emph{Isodata} so to show 
how the algorithm performs over.
%, which uses geometric information to improve the matching.  The 
%comparison  with \emph{Ratio} serves to illustrate the relative 
%performance of \emph{MM} and \emph{MMC} to a state of the art algorithm 
%without geometric constraints. \emph{Isodata} is included to illustrate 
%the performance of an algorithm relying on geometric constraints faced 
%with image pairs that might not have any overlap.  
The comparison was done using the \emph{Murals} dataset 
(Figure~\ref{fig:murals}), the \emph{Graf} set (Figure~\ref{fig:Graf}) 
from \cite{mikolajczyk2005performance}, and two 
images\footnote{100\_1942.jpg and 100\_1941.jpg} from the Gallagher 
dataset \cite{gallagher2008} (Figure~\ref{fig:comparemirror}).

@@

Figure~\ref{fig:result_graf} shows the results for 100 patch pairs 
generated from images 1 and 3 of the \emph{Graf} image set 
(cf.~Figure~\ref{fig:Graf}). The plot compares the accuracy of 
\emph{Ratio}, \emph{MM}, \emph{MMC}, and \emph{Isodata} as a function of 
the number of matches produced (which is achieved by varying thresholds 
over a certain range). The results show that \emph{MM} and \emph{MMC} 
consistently outperform \emph{Ratio}; \emph{MMC} generally lies 2-3 
percentage points above \emph{MM} when both are performing at optimal 
accuracy.  Although \emph{Isodata} exhibits a good performance on strict 
thresholds (small number of matches), that quickly diminishes when more 
matches are desired.

@@

In terms of computational complexity, \emph{Ratio}, \emph{MM}, and \emph{Isodata} can be 
implemented in $O(n\log n)$, where $n$ is the total number of feature 
points.  Our current \emph{MMC} implementation has a complexity of $O(n^2)$ due to the 
construction of a similarity matrix of the feature points. However, it is 
possible to approximate this in $O(n\log n)$ using search trees.

As for speed, Table~\ref{table:running_times} shows the running time of 
the four algorithms over 100 image pairs of $250\!\times\!250$ pixels.  
These numbers should be taken with a grain of salt, given that much of 
the code behind \emph{MMC} and \emph{Isodata} is implemented in Python, 
whereas \emph{MM} and \emph{Ratio} make use of OpenCV to execute 
computationally intensive operations in C++, which makes them much 
faster.

\begin{table}[htb]
\caption{Running times as tested on a Intel\textregistered\ Core\texttrademark\ i5-3550 CPU @ 
3.30~GHz with 8~GB memory.}
\label{table:running_times}
	\centering
%	\small
\begin{tabular}{r*{4}{c}}
\hline
	Algorithm: & \emph{Ratio} & \emph{MM} & \emph{MMC} & \emph{Isodata} 
	\\
	\noalign{\smallskip} 
	%
	Running time: & 20.94s & 18.23s & 883.99s & 555.33s \\
	\hline
\end{tabular}
\end{table}
%

@@

The algorithms show promising results when tested on the \emph{Murals} 
data set. The results on 900 image pairs show that \emph{MM} and 
\emph{MMC} generally outperform existing matching algorithms 
\emph{Ratio}, \emph{Isodata} and \emph{Spectral}. In addition \emph{MMC} 
outperforms \emph{MM}.  We show that this result generalizes to 
variations in viewpoint change by comparing \emph{MMC} to \emph{Ratio} 
over a set of image pairs with an increasing magnitude of perspective 
difference. We go on to apply the algorithms to a real life image case 
featuring occlusion and a slight viewpoint change and show that the 
performance of \emph{MM} and \emph{MMC} is consistent with the results 
on \emph{Murals}. To show that \emph{MM} or \emph{MMC}
can further improve the results of a geometrical method, we create 
\emph{Spectral-MMC} by combining \emph{Spectral} and \emph{MMC} and 
compare the three on the full \emph{Murals} dataset where it yields 
results better than both \emph{Spectral} and \emph{MMC}.

>>>>

		\State $matches_{init} \gets matches_{init} \cup \left(f_i, f_m\right)$

@@

\subsubsection{Experimental setup}
We compare the \emph{MM} and \emph{MMC} algorithms to \emph{Ratio} 
\cite{lowe2004sift} as well as \emph{Isomatch} \cite{das2008event} and 
\emph{Spectral} \cite{leordeanu2005spectral}.  \emph{Isomatch} and 
\emph{Spectral} uses geometric constraints, whereas \emph{Ratio} does 
not.

@@

\begin{algorithm}
\caption{Isodata Match (\emph{Isomatch})}
\label{alg-isodata}
%{\fontsize{10}{10}\selectfont
\begin{algorithmic}
\Require $im_1$ : image, $im_2$ : image, $t \in \mathbb{R}$
\State $matches_{init}\gets \varnothing$
\State $matches_{final}\gets \varnothing$
\State $features_1 \gets getFeatures(im_1)$
\State $features_2 \gets getFeatures(im_2)$
\ForAll{$f_i \in features_1$}
    \State $f_m,f_n \gets get2NNs(f_i, features_2)$
    \State $matches_{init} \gets \left(f_i, f_m\right)$
\EndFor
\State $partitions_1 \gets isodata(getPositions(features_1))$
\State $partitions_2 \gets isodata(getPositions(features_2))$
\State $P_{corr} \gets getMatchMat(partitions_1, partitions_2, 
matches_{init})$
\ForAll{$(i,j) \in indices(P)$}
    \If{$\left(P_{i,j} > 5\right) \wedge \left(P_{i,j} / sum(P_{i}) > 
    0.5\right)$}
        \State $matches_{final} \gets getMatches(matches_{init}, 
        partitions_1, partitions_2, i, j)$
    \EndIf
\EndFor
\Return $matches_{final}$
\end{algorithmic}
%}
\end{algorithm}

The \emph{Isomatch} algorithm (Algorithm~\ref{alg-isodata}) 
\cite{das2008event} is based on the idea that local feature points 
within a region in one image often matches to features within a 
corresponding region in the other image. To find these regions the 
algorithm uses the Isodata algorithm \cite{ball1965isodata} to perform 
unsupervised clustering on the feature points given their positions in 
the two images. The Isodata algorithm is similar to K-means with the 
addition that clusters are merged and split during the clustering 
process. The initial set of matches are then filtered based on the 
resulting regions. A match is kept only if it belongs to a region with 
more than 5 matches and where half of the matches link the same two 
regions. $P_{corr}$ in Algorithm~\ref{alg-isodata} is a matrix where 
each row corresponds to a region in $im_1$, each column corresponds to a 
region in $im_2$ and $P_{i,j}$ is the amount of matches going from 
region $i$ to region $j$.

To threshold the \emph{Isomatch} algorithm we filter the results using 
the ratio between first and second best match as in \emph{Ratio}, since 
\emph{Isomatch} itself doesn't readily admit any thresholding parameter.

\begin{algorithm}
\caption{Spectral Match (\emph{Spectral})}
\label{alg-spectral}
%{\fontsize{10}{10}\selectfont
\begin{algorithmic}
\Require $im_1$ : image, $im_2$ : image, $t \in \mathbb{R}$
\State $matches_{init}\gets \varnothing$
\State $matches_{final} \gets \varnothing$
\State $features_1 \gets getFeatures(im_1)$
\State $features_2 \gets getFeatures(im_2)$
\ForAll{$f_i \in features_1$}
	\State $f_m,f_n \gets get2NNs(f_i, features_2)$
	\State $matches_{init} \gets \left(f_i, f_m\right)$
\EndFor
\State $M \gets matrix(\left\vert matches_{init} \right\vert, \left\vert 
matches_{init} \right\vert)$
\ForAll{$m_i \in matches_{init}$}
	\ForAll{$m_j \in matches_{init}$}
		\If{$i = j$}
			\State $M \gets matchSimilarity(m_i)$
		\Else
			\State $M \gets affinity(m_i, m_j)$
		\EndIf
	\EndFor
\EndFor
\State $x^{*} \gets maximumEigenvector(M)$
\ForAll{$\left(m_i, x_i\right) \in \left(matches_{init}, x^{*}\right)$}
	\If{$x_i > t$}
		\State $matches_{final} \gets matchtes_{final} \cup m_i$
	\EndIf
\EndFor
\Return $matches_{final}$
\end{algorithmic}
%}
\end{algorithm}

The \emph{Spectral} algorithm (Algorithm~\ref{alg-spectral}) is a 
slightly modified version of the spectral matching algorithm proposed by
Leordeanu and Herbert \cite{leordeanu2005spectral}. It is based on the 
idea that the set of actual correspondences of feature points between 
two images should be pairwise consistent. That is, given two feature 
points that are geometrically close in one image, for most cases we 
would assume that the closest correspondences in another image would be 
similarly close. We want to pick out a set of matches $T \subset S$ 
where $S$ is the initial set of possible matches such that the matches 
in $T$ are as pairwise consistent as possible. If we let $x$ be a binary 
vector corresponding in length to the set of initial matches $S$, where 
$x_i = 1$ if $m_i \subset T$, then we can formulate a subset-score as:
\begin{equation*}
	S = \sum_{m_i, m_j \in T} M(m_i, m_j) = x^TMx
\end{equation*}
The solution $x^{*}$ that maximizes this equation is then $x^{*} = 
argmax(x^TMx)$. By relaxing the constraints on $x$ such that it can take
values in the range of $\left[0, 1\right]$ we can interpret $x_i$ as the
association of $m_i$ with the optimal subset of matches $T$. Because we 
only worry about the relative values of $x$ we can fix the norm of $x$ 
to $1$, in which case by Raleigh's ratio theorem, the $x^{*}$ that 
optimizes $x^TMx$ is the principal eigenvector of $M$.

Leordeanu and Herbert proposes a simple affinity function to map the 
pairwise closeness of two matches. Given a simple scene where no 
rotation and excessive affine transformation takes place, it can 
generally be assumed that the length and direction of the vector between
the two feature points constituting the first match corresponds to the 
length and direction of the vector between the two feature points 
constituting the second match. They formulated an affinity function as 
follows: \TODO: change this to real function
\begin{equation*}
	S = \sum_{m_i, m_j \in T} M(m_i, m_j) = x^TMx
\end{equation*}
There are a number of occasions where this assumption no longer holds 
true. A simple example would be matches between two images of the same 
scene where one image is rotated $180\degrees$. However for a lot of 
applications like panorama stitching
where we assume that images are generally not rotated a lot in between 
each other the assumption is accurate.

%, which uses geometric information to improve the matching.  The 
%comparison  with \emph{Ratio} serves to illustrate the relative 
%performance of \emph{MM} and \emph{MMC} to a state of the art algorithm 
%without geometric constraints. \emph{Isomatch} is included to 
%illustrate the performance of an algorithm relying on geometric 
%constraints faced with image pairs that might not have any overlap.  
The comparison was done using the \emph{Murals} dataset 
(Figure~\ref{fig:murals}), the \emph{Graf} set (Figure~\ref{fig:Graf}) 
from \cite{mikolajczyk2005performance}, and two 
images\footnote{100\_1942.jpg and 100\_1941.jpg} from the Gallagher 
dataset \cite{gallagher2008} (Figure~\ref{fig:comparemirror}).

@@

Figure~\ref{fig:result_graf} shows the results for 100 patch pairs 
generated from images 1 and 3 of the \emph{Graf} image set 
(cf.~Figure~\ref{fig:Graf}). The plot compares the accuracy of 
\emph{Ratio}, \emph{MM}, \emph{MMC}, and \emph{Isomatch} as a function of 
the number of matches produced (which is achieved by varying thresholds 
over a certain range). The results show that \emph{MM} and \emph{MMC} 
consistently outperform \emph{Ratio}; \emph{MMC} generally lies 2-3 
percentage points above \emph{MM} when both are performing at optimal 
accuracy.  Although \emph{Isomatch} exhibits a good performance on strict 
thresholds (small number of matches), that quickly diminishes when more 
matches are desired.

@@

In terms of computational complexity, \emph{Ratio}, \emph{MM}, and \emph{Isomatch} can be 
implemented in $O(n\log n)$, where $n$ is the total number of feature 
points.  Our current \emph{MMC} implementation has a complexity of $O(n^2)$ due to the 
construction of a similarity matrix of the feature points. However, it is 
possible to approximate this in $O(n\log n)$ using search trees.

As for speed, Table~\ref{table:running_times} shows the running time of 
the four algorithms over 100 image pairs of $250\!\times\!250$ pixels.  
These numbers should be taken with a grain of salt, given that much of 
the code behind \emph{MMC} and \emph{Isomatch} is implemented in Python, 
whereas \emph{MM} and \emph{Ratio} make use of OpenCV to execute 
computationally intensive operations in C++, which makes them much 
faster.

\begin{table}[htb]
\caption{Running times as tested on a Intel\textregistered\ Core\texttrademark\ i5-3550 CPU @ 
3.30~GHz with 8~GB memory.}
\label{table:running_times}
	\centering
%	\small
\begin{tabular}{r*{4}{c}}
\hline
	Algorithm: & \emph{Ratio} & \emph{MM} & \emph{MMC} & \emph{Isomatch} 
	\\
	\noalign{\smallskip} 
	%
	Running time: & 20.94s & 18.23s & 883.99s & 555.33s \\
	\hline
\end{tabular}
\end{table}
%

@@

The algorithms show promising results when tested on the \emph{Murals} 
data set. The results on 900 image pairs show that \emph{MM} and 
\emph{MMC} generally outperform existing matching algorithms 
\emph{Ratio}, \emph{Isomatch} and \emph{Spectral}. In addition \emph{MMC} 
outperforms \emph{MM}.  We show that this result generalizes to 
variations in viewpoint change by comparing \emph{MMC} to \emph{Ratio} 
over a set of image pairs with an increasing magnitude of perspective 
difference. We go on to apply the algorithms to a real life image case 
featuring occlusion and a slight viewpoint change and show that the 
performance of \emph{MM} and \emph{MMC} is consistent with the results 
on \emph{Murals}. To show that \emph{MM} or \emph{MMC}
can further improve the results of a geometrical method, we create 
\emph{Spectral-MMC} by combining \emph{Spectral} and \emph{MMC} and 
compare the three on the full \emph{Murals} dataset where it yields 
results better than both \emph{Spectral} and \emph{MMC}.
